The 2018 heatwave and its implications on ozone induced damage on vegetation in a subarctic climate

### 2020-06-05 VegOzone.tex
Included some of the citations which have been added.

{\bf TODO: Future climate and conditions
  \begin{itemize}
  \item[$\rightarrow$] increase in heatwaves and dry periods
  \item[$\rightarrow$] shift in start of growing season (into vicinity of ozone spring peak) \citep{GCB:Menzel2006,RS:Hogda2013}
  \item[$\rightarrow$] increase in tropospheric background ozone
  \end{itemize}
}

@@

\subsubsection{Surface ozone observations}
\label{subsubsec:ebas}
All surface ozone observations used in this study are available from the EBAS database operated by NILU (\href{http://ebas.nilu.no/}{ebas.nilu.no}). In Table~\ref{tab:ebas_obs}, we give an overview over the exact locations and time range of operational data taking. All stations with available long term observations are located on higher ground than Svanhovd which will result in higher surface ozone concentrations in comparison as ozone abundance, in general, decreases with decreasing altitude in the absence of local precursors. {\bf TODO: Citation see references in \citet{AB:Klingberg2009}!} {\bf TODO: The following has to be confirmed and elaborated on! Ane?} The vegetation at all sites is similar and consists mainly of pine forests, birch, and heath.

@@

From the northern Fennoscandic climatology, one can deduce several features. The maximum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, min}} peaks about $20\,\unit{days}$ earlier than the \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, and \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}}. With respect to the periodicity of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, it takes about $7.5\,\unit{months}$ from minimum concentration to maximum concentration, while from maximum to minimum it only takes about $4.5\,\unit{months}$. The maximum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}} is at $52\,\unit{ppb}$ in late April and the minimum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, min}} occurs at $15\,\unit{ppb}$ in the beginning of September. The steep rise of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}} in spring is in coincidence with increasing temperatures as well as increase in global irradiance. The decrease in \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}} begins about one month ahead of the estimated onset of the greening season. Based on data from SeNorge.no and the agricultural rule of thumb that the daily average temperature has to be above $5\,\unit{^\circ C}$ for $5\,\unit{days}$ consecutively, we find $\left<G_\mathrm{begin}\right> = (150 \pm 14)\,\unit{days of the year}$ at Svanvik (Appendix Fig~\ref{fig:greening_season_change_Svanvik}), which translates into the end of May.

>>>>

{\bf TODO: Future climate and conditions
  \begin{itemize}
  \item[$\rightarrow$] increase in heatwaves and dry periods
  \item[$\rightarrow$] shift in start of growing season (into vicinity of ozone spring peak) \citep{GCB:Menzel2006,RS:Hogda2013,IJB:Karlsen2007}
  \item[$\rightarrow$] increase in tropospheric background ozone cite WMO?
  \end{itemize}
}

@@

\subsubsection{Surface ozone observations}
\label{subsubsec:ebas}
All surface ozone observations used in this study are available from the EBAS database operated by NILU (\href{http://ebas.nilu.no/}{ebas.nilu.no}). In Table~\ref{tab:ebas_obs}, we give an overview over the exact locations and time range of operational data taking. All stations with available long term observations are located on higher ground than Svanhovd which will result in higher surface ozone concentrations in comparison as ozone abundance, in general, decreases with decreasing altitude in the absence of local precursors. {\bf TODO: Citation see references in \citet{AB:Klingberg2009}!} The periphery vegetation at all sites is similar and consists mainly of pine forests, birch, and heath. However, at close vicinity it varies substantially. Ozone measurements at Pallas are taken on top of mount Sammaltunturi $100\,\unit{m}$ above tree line (pine forest) and mainly consists of low vascular plants, mosses, and lichen \citep{BER:Hatakka2003}. At Svanhovd, the ozone measurements are taken amidst a meadow surrounded by agriculturally managed herbaceous plants and grass (c.c. Fig.\ref{fig:svanhovd_research_station}).

@@

From the northern Fennoscandic climatology, one can deduce several features. The maximum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, min}} peaks about $20\,\unit{days}$ earlier than the \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, and \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}}. With respect to the periodicity of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, it takes about $7.5\,\unit{months}$ from minimum concentration to maximum concentration, while from maximum to minimum it only takes about $4.5\,\unit{months}$. The maximum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}} is at $52\,\unit{ppb}$ in late April and the minimum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, min}} occurs at $15\,\unit{ppb}$ in the beginning of September. The steep rise of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}} in spring is in coincidence with increasing temperatures as well as increase in global irradiance. The decrease in \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}} begins about one month ahead of the estimated onset of the greening season. Based on data from SeNorge.no and the agricultural rule of thumb that the daily average temperature has to be above $5\,\unit{^\circ C}$ for $5\,\unit{days}$ consecutively, we find $\left<G_\mathrm{begin}\right> = (150 \pm 14)\,\unit{days of the year}$ at Svanvik (Appendix Fig~\ref{fig:greening_season_change_Svanvik}), which translates into the end of May. This finding is in line with more extensive studies \citep[][e.g.,]{IJB:Karlsen2007,AFM:Linderholm}. {\bf TODO: Rewrite how growing season has been calculated.}
